---
title: "ST558 Project 1: NHL Records & Stats Analysis"
author: "Lynn Huang"
date: "September 18, 2020"
output:
  rmarkdown::github_document
    toc: true
---
# Getting Started with NHL Records & NHL Stats APIs
This vignette examines how to access the NHL Records API data as hosted [here](https://gitlab.com/dword4/nhlapi/-/blob/master/records-api.md) and the NHL Stats API data as hosted [here](https://gitlab.com/dword4/nhlapi/-/blob/master/stats-api.md).  
All queries are prefixed with the base URL `https://records.nhl.com/site/api` or the base URL `https://statsapi.web.nhl.com/api/v1/` and are GET requests. The data is returned in JSON form.  
Information is available about items such as:  

* Franchises and Teams  
* Players, Drafts, Awards, and Milestones  
* Trophies, Playoffs, and Win-Loss Records  

## Required Packages
The following packages are required or highly recommended:  

* knitr  
* tidyverse 
* jsonlite  
* ggplot2

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
rm(list=ls())
library(knitr)
library(tidyverse)
library(jsonlite)
library(ggplot2)
# Include this in console execution ONLY to make your GitHub page:
#setwd("C:/Users/lynn/Documents/GitHub//ST558-Project-1/")
#rmarkdown::render(“Project.Rmd”, output_file=“README.md”)
#fid <- c(1, 2)
#fn <- c("Montréal Canadiens", "Montreal Wanderers")
```

## Functions
### Functions for NHL Records
Functions are available to contact the NHL Records API and return well-formatted, parsed data for:  

* Franchise: Team IDs, first and last seasons, and names for every team in the history of the NHL  
* Franchise Team Totals: Total Wins, Losses, and Ties for every franchise  
* Franchise Season Records: Season Wins, Losses, and Ties for every franchise  
* Franchise Goalie Records  
* Franchise Skater Records  

Filter Information:  

* Franchise ID. EX: franchId=1 returns information about the Montreal Canadiens  
  - The last 3 functions only accept franchiseID queries!  
  - The last 3 functions only accept single franchiseId queries!  
* Franchise Name. EX: franchName="Montréal Canadiens" returns information about the Montreal Canadiens  
  - Note that neither "Montréal" (nor "Montreal") nor "Canadiens" will work on their own  
* Multiple IDs or Multiple Names are permitted  

```{r recordsFunctions}
recordsBaseURL <- "https://records.nhl.com/site/api/"

# Function for /franchise (id, firstSeasonId, lastSeasonId, name)
getFranchise <- function(franchId=NULL, franchName=NULL) {
  value <- "franchise"
  URL <- paste0(recordsBaseURL, value)
  apiData <- jsonlite::fromJSON(URL)
  api.df <- as_tibble(apiData[[1]])
  # If available, filter by franchId
  if (!is.null(franchId)) {
    api.df <- api.df %>% filter(id %in% franchId)
  }
  # If available, filter by franchName (not the same as teamCommonName!)
  if (!is.null(franchName)) {
    # Split each franchName string into tokens and grab the last element only for teamCommonName
    fns <- rep("", length(franchName))
    for (i in 1:length(franchName)){
      fn.tokens <- strsplit(franchName[i], " ")
      fns[i] <- tail(fn.tokens[[1]], 1)
    }
    api.df <- api.df %>% filter(teamCommonName %in% fns)
  }
  return(api.df)
}
# getFranchise()
# getFranchise(franchId=fid)
# getFranchise(franchName=fn)
# getFranchise(franchId=fid, franchName=fn)

# Function for /franchise-team-totals (total stats)
getFranchiseTeamTotals <- function(franchId=NULL, franchName=NULL) {
  value <- "franchise-team-totals"
  URL <- paste0(recordsBaseURL, value)
  apiData <- jsonlite::fromJSON(URL)
  api.df <- as_tibble(apiData[[1]])
  # If available, filter by franchId
  if (!is.null(franchId)) {
    api.df <- api.df %>% filter(franchiseId %in% franchId)
  }
  # If available, filter by franchName (same as teamName)
  if (!is.null(franchName)) {
    api.df <- api.df %>% filter(teamName %in% franchName)
  }
  return(api.df)
}
# getFranchiseTeamTotals()
# getFranchiseTeamTotals(franchId=fid)
# getFranchiseTeamTotals(franchName=fn)
# getFranchiseTeamTotals(franchId=fid, franchName=fn)

# Function for /franchise-season-records (franchise-specific season records)
getFranchiseSeasonTotals <- function(franchId=NULL) {
  # Complain that you don't have franchId, which seems like it's required for this endpoint
  if (is.null(franchId)) {
    stop("I need a franchiseId!")
  }
  value <- "franchise-season-records?cayenneExp=franchiseId="
  URL <- paste0(recordsBaseURL, value, franchId)
  apiData <- jsonlite::fromJSON(URL)
  api.df <- as_tibble(apiData[[1]])
  return(api.df)
}
# fid=1
# getFranchiseSeasonTotals()
# getFranchiseSeasonTotals(franchId=fid)

# Function for /franchise-goalie-records (franchise-specific goalie records)
getFranchiseGoalieRecords <- function(franchId=NULL) {
  # Complain that you don't have franchId, which seems like it's required for this endpoint
  if (is.null(franchId)) {
    stop("I need a franchiseId!")
  }
  value <- "franchise-goalie-records?cayenneExp=franchiseId="
  URL <- paste0(recordsBaseURL, value, franchId)
  apiData <- jsonlite::fromJSON(URL)
  api.df <- as_tibble(apiData[[1]])
  return(api.df)
}
# getFranchiseGoalieRecords()
# getFranchiseGoalieRecords(franchId=fid)

# Function for /franchise-skater-records (franchise-specific skater records)
getFranchiseSkaterRecords <- function(franchId=NULL) {
  # Complain that you don't have franchId, which seems like it's required for this endpoint
  if (is.null(franchId)) {
    stop("I need a franchiseId!")
  }
  value <- "franchise-skater-records?cayenneExp=franchiseId="
  URL <- paste0(recordsBaseURL, value, franchId)
  apiData <- jsonlite::fromJSON(URL)
  api.df <- as_tibble(apiData[[1]])
  return(api.df)
}
# getFranchiseSkaterRecords()
# getFranchiseSkaterRecords(franchId=fid)

```

### Function for NHL Stats
A function is also available to contact the NHL Stats API and return well-formatted, parsed data for:  

* Teams: Team IDs, venues, divisions, etc. for every team in the history of the NHL  

Please note that you can filter the results returned specifically for these expanded modifiers. It is suggested that you filter, unless you want a big, ugly dataframe with all the possible information you could ever want to parse:  

* Team ID - This is different than franchiseId
  - teamId=1 is "New Jersey Devils" whereas franchiseId=1 is "Montréal Canadiens"  
  - Multiple teamId's can be provided as a vector of numbers  
* Season - The two-year identifier for an NHL season  
  - season="20192020" will get info for 2019-2020 season  
  - The API's default is to return the current season info (so, 2019-2020)  
* Team Roster - Roster of active players (if specified, for teamId and season)  
  - Return: A dataframe or list of dataframes if multiple teamId's specified  
* Person Names - Just player names (if provided, for teamId)  
  - Return: A dataframe of team name and info (each row)  
* Team Schedule Next - Details of upcoming games (if provided, for teamId)  
  - Not all teams have upcoming games, try teamId=14 (worked as of 9-18-2020)  
  - Return: A dataframe or list of dataframes if multiple teamId's specified  
* Team Schedule Previous - Details of previous games (if provided, for teamId)  
  - Return: A dataframe or list of dataframes if multiple teamId's specified  
* Team Stats - Team win/loss/tie type of stats (if provided, for teamId)  
  - Return: A dataframe (of dataframes/lists)
* statsSingleSeasonPlayoffs (ignored due to unclear documentation in the NHL Stats API)  

You may only select teamRoster OR personNames or teamSchedNext or teamSchedPrev or teamStats, not a combination of them.  
```{r statsFunction}
statsBaseURL <- "https://statsapi.web.nhl.com/api/v1/teams"

# First, a helper function to append teamId and season modifiers to the other 5 stats modifiers
getStatsHelper <- function(URL, teamId=NULL, season=NULL) {
  # Update with teamId and season data if applicable
  if (!is.null(teamId)) {
    tids <- as.character(teamId[1])
    if (length(teamId) > 1) {
      for (t in 2:length(teamId)) {
        tids <- paste0(tids, ",", teamId[t])
      }
    }
    URL <- paste0(URL, "&teamId=", tids)
  }
  if (!is.null(season)) {
    URL <- paste0(URL, "&season=", season)
  }
  return(URL)
}

# Second, the main function that actually gets API data
getStats <- function(teamId=NULL, season=NULL, teamRoster=FALSE, personNames=FALSE, 
                    teamSchedNext=FALSE, teamSchedPrev=FALSE, teamStats=FALSE) {
  # If only 1 teamId is provided (all modifiers are FALSE), don't use any modifiers
  if (!teamRoster & !personNames & !teamSchedNext & !teamSchedPrev & !teamStats & length(teamId)==1) {
    URL <- paste0(statsBaseURL, "/", teamId)
    apiData <- jsonlite::fromJSON(URL)
    api.df <- as_tibble(apiData[[2]])
    return(api.df)
  }
  
  # If any modifiers are used with teamId and season, update URL before grabbing data from API
  if (teamRoster) {
    URL <- paste0(statsBaseURL, "?expand=team.roster")
    URL <- getStatsHelper(URL, teamId, season)
    apiData <- jsonlite::fromJSON(URL)
    api.df <- apiData[[2]]
    rosters <- api.df$roster$roster
    return(rosters)
  }
  else if (personNames) {
    URL <- paste0(statsBaseURL, "?expand=person.names")
    URL <- getStatsHelper(URL, teamId, season)
    apiData <- jsonlite::fromJSON(URL)
    api.df <- as_tibble(apiData[[2]])
    return(api.df)
    }
  else if (teamSchedNext) {
    URL <- paste0(statsBaseURL, "?expand=team.schedule.next")
    URL <- getStatsHelper(URL, teamId, season)
    apiData <- jsonlite::fromJSON(URL)
    api.df <- as_tibble(apiData[[2]])
    nextGames <- api.df$nextGameSchedule$dates
    return(nextGames)
    }
  else if (teamSchedPrev) {
    URL <- paste0(statsBaseURL, "?expand=team.schedule.previous")
    URL <- getStatsHelper(URL, teamId, season)
    apiData <- jsonlite::fromJSON(URL)
    api.df <- as_tibble(apiData[[2]])
    prevGames <- api.df$previousGameSchedule$dates
    return(prevGames)
    }
  else if (teamStats) {
    URL <- paste0(statsBaseURL, "?expand=team.stats")
    URL <- getStatsHelper(URL, teamId, season)
    apiData <- jsonlite::fromJSON(URL)
    api.df <- apiData[[2]]
    return(api.df)
    }
  else {
    URL <- statsBaseURL
    apiData <- jsonlite::fromJSON(URL)
    api.df <- apiData[[2]]
    return(api.df)
  }
}
# rosters12 <- getStats(teamId=c(1,2), season="20192020", teamRoster=TRUE)
# getStats(teamId=c(1,2), season="20192020", personNames=TRUE)
# getStats(teamId=c(1,14), season="20192020", teamSchedNext=TRUE)
# getStats(teamId=c(1,14), season="20192020", teamSchedPrev=TRUE)
# getStats(teamId=c(1,2), season="20192020", teamStats=TRUE)
```
### Wrapper Function
Here is a on-stop-shop function to access any endpoint from the previous 6 functions. Input a funcType number from 1 to 6 for particular function calls, and provide optional arguments:  

1. Franchise (franchId, franchName)  
2. Franchise Goalie Records (franchId)  
3. Franchise Season Totals (franchId)  
4. Franchise Skater Records (franchId)  
5. Franchise Team Totals (franchId, franchName)  
6. Team and Season Level Stats (teamId, season, teamRoster, personNames, teamSchedNext, teamSchedPrev, teamStats)  
```{r Wrapper}
getNHLData <- function(funcType=c("franchise", "goalie", "season", "skater", "team", "stats"), ...) {
  switch(EXPR=funcType,
         franchise = getFranchise(...),
         goalie = getFranchiseGoalieRecords(...),
         season = getFranchiseSeasonTotals(...),
         skater = getFranchiseSkaterRecords(...),
         team = getFranchiseTeamTotals(...),
         stats = getStats(...))
}
# Testing
# f <- getNHLData("franchise", franchId=14)
# fg <- getNHLData("goalie", franchId=14)
# fs <- getNHLData("season", franchId=14)
# fk <- getNHLData("skater", franchId=14)
# ft <- getNHLData("team", franchName="Los Angeles Kings")
# s <- getNHLData("stats", teamId=26, teamRoster=TRUE)

```

## Exploratory Data Analysis
I'm originally from Los Angeles, so let's examine some of the data for the Los Angeles Kings. I still remember when they won the Stanley Cup in 2012 and 2014, and thinking how odd that we could win hockey tournaments during the longest drought in California (from 2011-2016).  
```{r Kings Info}
f <- getNHLData("franchise", franchId=14)
```
Their franchiseId is `r f$id`, teamId is `r f$mostRecentTeamId`, and they first started playing in the `r f$firstSeasonId` and are still playing.  

### Join
We can join the information about Franchise Goalie Records and Franchise Skater Records (both from the Records API) with the team roster from the Stats API to get stats on the current team roster. Both of these are left joins, where we wish to keep all information in the first table (team roster from Stats API) and add information from the second table (either goalie records or skater records).   
```{r}
fg <- getNHLData("goalie", franchId=14)
fk <- getNHLData("skater", franchId=14)
s <- getNHLData("stats", teamId=26, teamRoster=TRUE)
# Stats API function returns a list with dataframes, so make it into a nice dataframe
s <- cbind(s[[1]]$person, s[[1]]$jerseyNumber, s[[1]]$position)

# Get active players only
fg.act <- fg %>% filter(activePlayer=TRUE) %>% select(-id, -activePlayer, -franchiseId, -franchiseName,)
fk.act <- fk %>% filter(activePlayer=TRUE) %>% select(-activePlayer, -franchiseId, -franchiseName)
s.act <- s %>% filter(activePlayer=TRUE) %>% select(-link)

# Join by playerId (called id in s.act) to get goalie and skater info about current team roster
roster.Info.g <- left_join(s.act %>% filter(type=="Goalie"), fg.act, by=c("id"="playerId"))
roster.Info.fk <- left_join(s.act %>% filter(type!="Goalie"), fk.act, by=c("id"="playerId"))
roster.Info.g
roster.Info.fk
```
It appears there's no information about LA Kings' goalie Calvin Petersen. Some quick Googling shows he actually plays mostly in the AHL (American Hockey League) for the Ontario Reign team, which is the AHL affiliate of the LA Kings. This API is based on NHL data, and thus would miss out on Calvin Petersen's info.  
Otherwise, the skater info seems to have populated well, with no such missing player info.  

### Create New Variables
I wonder if there is a correlation between the number of games played and the percentage of wins for goalies and skaters? It seems reasonable to think a team would play the best players most often.  
I also wonder if having played as a rookie might influence a player's win percentage, so let's make a categorial variable for that.  
```{r}
# For goalies, let's calculate win percentage, and flag if someone is a rookie
fg.summStats <- fg %>% select(firstName, lastName, wins, gamesPlayed, rookieGamesPlayed) %>%
  mutate(winPct = round(wins/gamesPlayed, 2)) %>% 
  mutate(rookieFlag=factor(ifelse(is.na(rookieGamesPlayed), 0, 1)))

# For skaters, let's calculate goals per game
fk.summStats <- fk %>% select(firstName, lastName, goals, gamesPlayed, rookiePoints) %>% 
  mutate(gpg = round(goals/gamesPlayed, 2)) %>% 
  mutate(rookieFlag=factor(ifelse(is.na(rookiePoints), 0, 1)))
```

### Goalies: Contingency Tables, Numerical Summaries, Graphical Summaries
Let's examine if starting as a rookie influenced win percent in goalies or goals per game in skaters.  
About 16.3% of the LA Kings goalies once played as rookies.  
Looking at the boxplot, we see that goalies who started as rookies tended to win more games. Perhaps a team only invests in a goalie as a rookie if they show unusual promise or high performance.  
```{r Goalies}
# 1-Way Table
(t1 <- table(fg.summStats$rookieFlag))
t1[2]/(t1[1] + t1[2])

# Win Percent by Rookie Flag
# Built-in summary() table returns something ugly, so we write our own function
make_table <- function(x, cn){
  min <- min(x)
  q1 <- quantile(x, probs=0.25)
  med <- median(x)
  mean <- mean(x)
  q3 <- quantile(x, probs=0.75)
  max <- max(x)
  tab <- round(rbind(min, q1, med, mean, q3, max), 2)
  rownames(tab) <- c("Min.", "1st Q.", "Median", "Mean", "3rd Q.", "Max.")
  colnames(tab) <- cn
  kable(tab)
}

# Let's look at the 5-number summaries for rookies vs. non-rookies
fg.winPct1 <- unlist(fg.summStats %>% filter(rookieFlag==1) %>% select(winPct))
make_table(fg.winPct1, cn="Started as Rookie")
fg.winPct0 <- unlist(fg.summStats %>% filter(rookieFlag==0) %>% select(winPct))
make_table(fg.winPct0, cn="Didn't Start as Rookie")

# Compare using box plots side-by-side
g <- ggplot(data=fg.summStats, aes(x=rookieFlag, y=winPct)) + 
  labs(title="Goalie Win Percent Depending On If They Started as Rookies", 
       x="Started as Rookie", 
       y="Win Percent")
g + geom_boxplot(fill="grey") + scale_x_discrete(labels=c("No", "Yes"))
```

### Skaters: Contingency Tables, Numerical Summaries, Graphical Summaries
About 44.0% of the LA Kings skaters once played as rookies. Quite different from the goalies!
Perhaps due to this greater likelihood of accepting skaters as rookies, there's not much difference in their goals per game (see boxplot).  
```{r Skaters}
# 1-Way Table
(t3 <- table(fk.summStats$rookieFlag))
t3[2]/(t3[1] + t3[2])

# Goals Per Game by Rookie Flag
# Let's look at the 5-number summaries for rookies vs. non-rookies
fk.gpg1 <- unlist(fk.summStats %>% filter(rookieFlag==1) %>% select(gpg))
make_table(fk.gpg1, cn="Started as Rookie")
fk.gpg0 <- unlist(fk.summStats %>% filter(rookieFlag==0) %>% select(gpg))
make_table(fk.gpg0, cn="Didn't Start as Rookie")

# Compare using box plots side-by-side
g2 <- ggplot(data=fk.summStats, aes(x=rookieFlag, y=gpg)) + 
   labs(title="Skater Goals Per Game Depending On If They Started as Rookies", 
        x="Started as Rookie", 
        y="Games Per Goal")
g2 + geom_boxplot(fill="grey") + scale_x_discrete(labels=c("No", "Yes"))
```

### Franchises & Teams: More Plots
Let's take a break from the LA Kings and look at all teams as a group.  
```{r Franchises & Teams}
# Get data for last season of games (splits: stat and rankings)
ts <- getNHLData("stats", teamStats=TRUE, season="20192020")
attributes(ts)

# Pull out teamId and teamName.
n = nrow(ts)
# For each of the 31 teams that are currently active, pull out info about the division and splits (df)
# You'll need to query for each team individually
id = as.vector(ts$id)
name = as.vector(ts$name)
div = rep(NA, n)
conf = rep(NA, n)
gamesPlayed = rep(NA, n)
wins = rep(NA, n)
losses = rep(NA, n)
pts = rep(NA, n)
goalsPerGame = rep(NA, n)
goalsAgainstPerGame = rep(NA, n)
for (i in 1:n) {
  team <- getNHLData("stats", teamStats=TRUE, teamId=tids[i])
  div[i]                 <- team$division$name
  conf[i]                <- team$conference$name
  gamesPlayed[i]         <- team$teamStats[[1]]$splits[[1]]$stat$gamesPlayed[1]
  wins[i]                <- team$teamStats[[1]]$splits[[1]]$stat$wins[1]
  losses[i]              <- team$teamStats[[1]]$splits[[1]]$stat$losses[1]
  pts[i]                 <- team$teamStats[[1]]$splits[[1]]$stat$pts[1]
  goalsPerGame[i]        <- team$teamStats[[1]]$splits[[1]]$stat$goalsPerGame[1]
  goalsAgainstPerGame[i] <- team$teamStats[[1]]$splits[[1]]$stat$goalsAgainstPerGame[1]
}

# Put it all together in a new dataframe
teams.df <- cbind(id, gamesPlayed, as.numeric(wins), as.numeric(losses), as.numeric(pts),
                  as.numeric(goalsPerGame), as.numeric(goalsAgainstPerGame))
teams.df <- as.data.frame(teams.df)
colnames(teams.df) <- c("id", "gamesPlayed", "wins", "losses", "pts", "goalsPerGame",
                        "goalsAgainstPerGame")
teams.df <- cbind(teams.df, name, div, conf)
```

How many teams are in each division? Pretty evenly split, as we expected.  
```{r Bar Plot}
g3 <- ggplot(data=teams.df, aes(x=div)) + labs(x="Division")
g3 + geom_bar()
```


Let's plot goals scored vs goals allowed to see where offense and defense intersect.
Teams in the bottom right corner perform the best, scoring more goals per game while keeping goals allowed (by opponents) per game low. There's no obvious patterns by division, but we can look at the outliers a bit more (both are Atlantic teams)
```{r Scatterplot}
g4 <- ggplot(teams.df, aes(x=goalsPerGame, y=goalsAgainstPerGame)) + 
  labs(title="Goals Allowed vs. Goals Scored", x="Goals Scored Per Game", y="Goals Allowed Per Game")
g4 + geom_point(aes(color=factor(div))) +
  scale_color_manual(name="Division", values=c("green", "orange", "red", "blue"))
```

Looks like the Detroit Red Wings performed unusually poor (allowing more goals per game while scoring few goals per game) and the Boston Bruins performed unusually well (scoring more goals per game while allowing fewer goals per game).  
```{r Scatterplot 2}
atlanticTeams.df <- teams.df %>% filter(div=="Atlantic")
g5 <- ggplot(atlanticTeams.df, aes(x=goalsPerGame, y=goalsAgainstPerGame)) + 
  labs(title="Goals Allowed vs. Goals Scored", x="Goals Scored Per Game", y="Goals Allowed Per Game")
g5 + geom_point() + geom_text(aes(label=name))
```

Let's look at the overall distribution of player scoring for the Red Wings and Bruins. Is there a noticeable difference?  
There appear to be more high-scoring members on the Bruins than the Red Wings.  
```{r Histogram}
# Red Wings teamId=17, Bruins teamId=6 (teams.df), which corresponds to franchId=12 and 6 (dataframe f)
redwings.df <- getNHLData(funcType="skater", franchId=12) %>% filter(activePlayer==TRUE)
bruins.df <- getNHLData(funcType="skater", franchId=6) %>% filter(activePlayer==TRUE)
# Make sure to filter by active members only!

g6 <- ggplot(data=redwings.df, aes(x=points)) + 
  labs(title="Points per Skater in the Detroit Red Wings")
g6 + geom_histogram(binwidth=50, color="red", fill="red")

g7 <- ggplot(data=bruins.df, aes(x=points)) + 
  labs(title="Points per Skater in the Boston Bruins")
g7 + geom_histogram(binwidth=50, color="gold", fill="gold")
```